Individual heterogeneity in functional topography of the default-mode network in medial prefrontal cortex

### 2019-03-28 Manuscript.Rmd
halfway through the results section

All network analyses were performed using the igraph package [@G.2006] in R 3.4.1 [@RTeam2018]. To establish each subject's network, we selected all the vertices contained within the mPFC/PCC search space (n = 4,801 per subject; mPFC = 2854, PCC = 1947) and computed the Pearson correlation of the time series for every pair of vertices. This produced a weighted network for each subject, in which the nodes were surface vertices and the edges were the correlations among them. Next, we applied Fisher's r to z transformation, and performed a two-sided significance test to identify strong correlations. The resulting p-values were corrected for multiple comparisons (FDR < 0.05), and edges with non-significant correlations were turned to 0 (all other edges retained their respective z-values). *TODO: recompute corrCounts so it also stores the # of pos/neg correlations turned to 0, as well as the number of TRs per subject (add to fMRI deets)*. Next, we took the exponential of these values so that all weights were positive while maintaining the ordinal ranks of the original correlation distribution. Values of 1 of this transformed distribution were returned to 0 in order to avoid the inclusion of these edges in the network analyses. The proportion of remaining edges (i.e. network density) was high for all individuals(mean = `r round(mdAll, digits = 2)`, SD = `r round(sddAll, digits = 3)`). We generated and analyzed network weight matrices at four levels: 1) for each run separately (1200 TRs each); 2) for the concatenated time series for the two runs on each day (2400 TRs total); 3) for each subject's full dataset (4800 TRs total); and 4) on each step of a sliding window analysis (see Partition Evaluation for more details).

@@

In order to evaluate the convergent validity of the resulting partitionings across community-detection methods, we also estimated these communities using the more traditional approach of modularity maximization, based on the algorithm from Clauset et al. [-@Clauset2004]. *(TODO: rephrase) This method maximizes a quality function that compares the strength of the connection between any two vertices against a null model of their probability of being randomly connected to any other vertex in the network. The method heuristically iterates through many possible combinations of vertices, and selects the clustering that maximizes the network's modularity.* Unlike SP, modularity can fractionate a network into more than two communities. We assumed that if the partitionings from the bounded and unbounded community detection methods showed high levels of agreement, we could gain confidence in that we were indeed fractioning the search space into an adequate number of communities. We also diagnosed our choice of two communities by reporting the rank (from lowest to largest) of the first eigenvalue produced by SP to be non-zero. This eigenvalue serves as an estimator of the optimal number of communities in a network [@Chung1999], such that a rank of two in our partitions justifies limiting the number of communities to two.

@@

Before evaluating the degree of generazibility of this topological pattern across individuals, we examined the validity of these partitions by comparing them to patterns derived with a modularity algorithm [@Clauset2004]. Modularity seeks to find the pattern of communities that maximizes a quality function that compares the connectivity strength of any two nodes with a null probability model. Since modularity is not bound to a desired number of communities, it was capable of finding more than two in our datset. We quantified the cross-method agreement with the adjusted rand index (ARI). ARI measures the proportion of pairs of nodes in a network that were either clustered together or separately in both partitions, over all possible affiliation changes. This measure has the benefit of being agnostic to labeling schemes, as well as controlling for the expectation that the clusters were randomly produced [@Hubert1985]. This means that positive ARI values compare favorably against chance clustering.

@@

This comparison showed that both clustering methods were in high agreement across subjects (mean = `r round(mean(RI_all_within$Index), digits = 2)`, SD = `r round(sd(RI_all_within$Index), digits = 2)`). Modularity showed a tendency to produce additional communities (median = `r median(nModularity)`, max = `r max(nModularity)`). However, these communities encompassed a small number of vertices (mean = `r mean(modularityCounts_extra, na.rm = T)`, IQR = `r IQR(modularityCounts_extra, na.rm = T)`) compared to the principal two (mean = `r mean(modularityCounts_two, na.rm = T)`, IQR = `r IQR(modularityCounts_two, na.rm = T)`), suggesting that the overall network is composed of two main communities. In addition, the estimation of the optimal number of cuts based on the rank of the first non-zero eigenvalue (see "Partition evaluation" for details) showed that `r prcntDissection` percent of the individual networks are best partitioned into two communities.

Next, we examined how similar the partitions were across subjects by computing the ARI between every pair of subjects, and found that there was moderate agreement overall (mean = `r round(mean(RI_all_between_lowtri), digits = 2)`, SD = `r round(sd(RI_all_between_lowtri), digits = 2)`). However, qualitative inspection of the brain community topologies showed good alignment for PCC, with a topologically similar mPFC pattern that shifted topographically across subjects. To quantify this heterogeneity in mPFC, we performed the same analysis for each region separately (Figure 3). Indeed, this analysis showed that the functional topology of PCC is better maintained across subjects (mean = `r round(mean(testPCC), digits = 2)`, SD = `r round(sd(testPCC), digits = 2)`) than mPFC (mean = `r round(mean(testPFC), digits = 2)`, SD = `r round(sd(testPFC), digits = 2)`)(paired permutation, *p* = `r Overall_PCCvsPFC_permTest$Pval`; Cohen's D = `r round(Overall_PCCvsPFC_ES, digits = 2)`).

@@

In order to estimate the degree of stability of these partitions in time, we performed a sliding window analysis on each subject's full time series (20 min windows shifting at 1 min), comparing the community arrangement derived from each window with the partition computed using the whole time series (Figure 4).

@@

This analysis showed that the mean ARI along each subject's time series was high for both PCC (mean = `r round(mean(SC_PCCPFC_summary$Index[SC_PCCPFC_summary$Region=="PCC"]), digits = 2)`; SD = `r round(sd(SC_PCCPFC_summary$Index[SC_PCCPFC_summary$Region=="PCC"]), digits = 2)`) and mPFC (mean = `r round(mean(SC_PCCPFC_summary$Index[SC_PCCPFC_summary$Region=="mPFC"]), digits = 2)`; SD = `r round(sd(SC_PCCPFC_summary$Index[SC_PCCPFC_summary$Region=="mPFC"]), digits = 2)`), with significantly higher ARI for PCC (paired permutation, *p* = `r SW_PCCvPFC_permuteTest$Pval`; Cohen's D = `r round(SW_PCCvPFC_ES, digits = 2)`). These results suggest that the each individual's DMN distribution is cosiderably stable in time.

There is precedence in thhe literature that the magnitude (and not just the sign) of the Fiedler vector values conveys important information about the role of each node in the network [@Gkantsidis2003; @Tian2018]. Therefore, we asked whether the magnitude of these eigenvector values was associated with the stability of nodes in time. Specifically, we examined whether Fiedler vector values could predict the proportion of DMN affiliations per node by performing a logistic mixed effects model (Figure 5). The model yielded a positive significant relationship between these features ($\beta$ = `r round(SC_FVvPDMN_modelSummary$coefficients[2,1], digits = 2)`, SE = `r round(SC_FVvPDMN_modelSummary$coefficients[2,2], digits = 2)`, *p* < 0.001), meaning that vertices with higher absolute Fiedler vector values were more persistent in their relationship with their corresponding community in time. We refer to this proportion of affiliations over time as the stability of a given node.

@@

These analyses show that the original pattern identified with the full dataset is stable in time, and that there is potential value in thresholding the Fiedler vector as a means to capture idiosyncratic and reliable DMN and non-DMN vertices, even in relatively short resting fMRI sessions (i.e. less than 20 mins). We therefore thresholded each subject's Fiedler vector to produce these refined maps. We estimated this threshold by selecting the smallest absolute Fiedler vector value that yielded a stability of 95%, for positive (mean = `r threshDescriptives[3]`, SD = `r threshDescriptives[4]`) and negative (mean = `r threshDescriptives[1]`, SD = `r threshDescriptives[2]`) values separately. Individuals without such stable nodes (n = `r nUnstable`) were not thresholded, but their partitions were not discarded. Vertices under these thresholds were set to zero (instead of removed) so that the dimensionality of the data was unaffected, and 0.5 in the binarized maps so that they wouldn't bias the calculation of averages. Figure 6 (left) shows the updated partition for the one showed in figure 2. The maps used in all subsequent analyses were thresholded by this individualized criterion.

>>>>

All network analyses were performed using the igraph package [@G.2006] in R 3.4.1 [@RTeam2018]. To establish each subject's network, we selected all the vertices contained within the mPFC/PCC search space (n = 4,801 per subject; mPFC = 2854, PCC = 1947) and computed the Pearson correlation of the time series for every pair of vertices. This produced a weighted network for each subject, in which the nodes were surface vertices and the edges were the correlations among them. Next, we applied Fisher's r to z transformation, and performed a two-sided significance test to identify strong correlations. The resulting p-values were corrected for multiple comparisons (FDR < 0.05), and edges with non-significant correlations were turned to 0 (all other edges retained their respective z-values). **TODO: recompute corrCounts so it also stores the # of pos/neg correlations turned to 0, as well as the number of TRs per subject (add to fMRI deets)**. Next, we took the exponential of these values so that all weights were positive while maintaining the ordinal ranks of the original correlation distribution. Values of 1 of this transformed distribution were returned to 0 in order to avoid the inclusion of these edges in the network analyses. The proportion of remaining edges (i.e. network density) was high for all individuals(mean = `r round(mdAll, digits = 2)`, SD = `r round(sddAll, digits = 3)`). We generated and analyzed network weight matrices at four levels: 1) for each run separately (1200 TRs each); 2) for the concatenated time series for the two runs on each day (2400 TRs total); 3) for each subject's full dataset (4800 TRs total); and 4) on each step of a sliding window analysis (see Partition Evaluation for more details).

@@

In order to evaluate the convergent validity of the resulting partitionings across community-detection methods, we also estimated these communities using the more traditional approach of modularity maximization, based on the algorithm from Clauset et al. [-@Clauset2004]. **(TODO: rephrase) This method maximizes a quality function that compares the strength of the connection between any two vertices against a null model of their probability of being randomly connected to any other vertex in the network. The method heuristically iterates through many possible combinations of vertices, and selects the clustering that maximizes the network's modularity.** Unlike SP, modularity can fractionate a network into more than two communities. We assumed that if the partitionings from the bounded and unbounded community detection methods showed high levels of agreement, we could gain confidence in that we were indeed fractioning the search space into an adequate number of communities. We also diagnosed our choice of two communities by reporting the rank (from lowest to largest) of the first eigenvalue produced by SP to be non-zero. This eigenvalue serves as an estimator of the optimal number of communities in a network [@Chung1999], such that a rank of two in our partitions justifies limiting the number of communities to two.

@@

Before evaluating the degree of generazibility of this topographic pattern across individuals, we examined the validity of these partitionings by comparing them to results from an alterntive community detection algorithm, modularity maximization [@Clauset2004]. Modularity seeks to find the set of communities that maximizes a quality function that compares the connectivity strength of any two nodes with a null probability model. Since modularity is not bound to a desired number of communities, it was capable of finding more than two in our data set. We quantified the cross-method agreement in terms of the Adjusted Rand Index (ARI; see methods), which measures the proportion of node pairs of nodes in a network that were either clustered together or separately in both partitionings, while being agnostic to labeling schemes and controlling for chance clustering.

@@

This comparison showed that both clustering methods were in high agreement across subjects (mean ARI = `r round(mean(RI_all_within$Index), digits = 2)`, SD = `r round(sd(RI_all_within$Index), digits = 2)`). Modularity showed a tendency to produce additional communities (median = `r median(nModularity)`, max = `r max(nModularity)`). However, these communities encompassed a **(TBD: small number of vertices (mean = `r mean(modularityCounts_extra, na.rm = T)`, IQR = `r IQR(modularityCounts_extra, na.rm = T)`) compared to the principal two (mean = `r mean(modularityCounts_two, na.rm = T)`, IQR = `r IQR(modularityCounts_two, na.rm = T)`)**, suggesting that the overall network is composed of two main communities. In addition, the estimation of the optimal number of cuts based on the rank of the first non-zero eigenvalue (see "Partition evaluation" for details) showed that `r prcntDissection` percent of the individual networks were best partitioned into two communities.

Next, we examined how similar the partitions were across subjects by computing the ARI between every pair of subjects, and found that there was moderate above-chance agreement overall (mean = `r round(mean(RI_all_between_lowtri), digits = 2)`, SD = `r round(sd(RI_all_between_lowtri), digits = 2)`). Qualitative inspection of the brain community topologies showed good alignment for PCC, with a topologically similar mPFC pattern that shifted topographically across subjects. To quantify this heterogeneity in mPFC, we performed the same analysis for each region separately (Figure 3). Indeed, this analysis showed that the functional topography of PCC was better maintained across subjects (mean = `r round(mean(testPCC), digits = 2)`, SD = `r round(sd(testPCC), digits = 2)`) than mPFC (mean = `r round(mean(testPFC), digits = 2)`, SD = `r round(sd(testPFC), digits = 2)`)(paired permutation, *p* `r ifelse(Overall_PCCvsPFC_permTest$Pval == 0, "< 0.001", paste("=", Overall_PCCvsPFC_permTest$Pval))`; Cohen's D = `r round(Overall_PCCvsPFC_ES, digits = 2)`).

@@

In order to estimate the degree of stability of these partitions in time, we performed a sliding window analysis on each subject's full time series (20 min windows shifting by 1 min), comparing the partitioning derived from each window with the partitioning computed using the entire time series (Figure 4).

@@

**This analysis showed that the mean ARI along each subject's time series was high for both PCC (mean = `r round(mean(SC_PCCPFC_summary$Index[SC_PCCPFC_summary$Region=="PCC"]), digits = 2)`; SD = `r round(sd(SC_PCCPFC_summary$Index[SC_PCCPFC_summary$Region=="PCC"]), digits = 2)`) and mPFC (mean = `r round(mean(SC_PCCPFC_summary$Index[SC_PCCPFC_summary$Region=="mPFC"]), digits = 2)`; SD = `r round(sd(SC_PCCPFC_summary$Index[SC_PCCPFC_summary$Region=="mPFC"]), digits = 2)`), with significantly higher ARI for PCC (paired permutation, *p* = `r SW_PCCvPFC_permuteTest$Pval`; Cohen's D = `r round(SW_PCCvPFC_ES, digits = 2)`). These results suggest that some nodes were considerably variable in time, despite overall stability in the community arrangement.**

There is precedent in thhe literature for the idea that the magnitude (and not just the sign) of the Fiedler vector values conveys important information about the role of each node in the network [@Gkantsidis2003; @Tian2018]. Therefore, we asked whether the magnitude of these eigenvector values was associated with the stability of nodes in time. Specifically, we estimated the proportion of DMN affiliations per node as a function of Fiedler vector values, using a logistic mixed effects model (Figure 4). The model yielded a positive significant relationship between these features ($\beta$ = `r round(SC_FVvPDMN_modelSummary$coefficients[2,1], digits = 2)`, SE = `r round(SC_FVvPDMN_modelSummary$coefficients[2,2], digits = 2)`, *p* < 0.001), meaning that vertices with **higher absolute Fiedler vector values** were more persistent in their relationship with their corresponding community over time. We refer to this proportion of affiliations over time as the stability of a given node.

@@

These analyses show that the original pattern identified with the full dataset is stable in time, and that there is potential value in thresholding the Fiedler vector as a means to identify idiosyncratic and reliable DMN and non-DMN vertices. We therefore thresholded each subject's Fiedler vector to produce these refined maps. We estimated this threshold by selecting the smallest absolute Fiedler vector value that yielded a stability of 95%, for positive (mean = `r threshDescriptives[3]`, SD = `r threshDescriptives[4]`) and negative (mean = `r threshDescriptives[1]`, SD = `r threshDescriptives[2]`) values separately. Individuals without such stable nodes (n = `r nUnstable`) were not thresholded, and were included in the subsequent analyses in unthresholded form. Sub-threshold vertices were set to zero in Fiedler vector maps and 0.5 in the binarized maps so that they would not bias the calculation of averages. Figure 5 shows the thresholded partitioning for the same individual shown in Figure 2. The maps used in all subsequent analyses were thresholded by this individualized criterion.
